\LARGE{\bf Parallelizing a Network Simulator}

### 2009-03-16 doc/disc.tex
I am good with it
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1158 ce7aba0d-9f81-40a6-a495-cecae84b0515

\subsection{Static Network Partition}
The current implementation divides the network into equal partitions among the threads and seeks to minimize computation. This organization works well when the traffic pattern is balanced across the partitions. However, hotspot traffic patterns which concentrate the packets into a particular section of the network will cause imbalances in the amount of work handled by each thread. Even with the shared channel semaphore and the removal of the cycle barrier, the lightly loaded threads will almost always be stalled waiting for the hotspot threads. Fortunately, common network traffic patterns are balanced and network routing algorithm are usually designed to reduce hotspots and bottlenecks in the network. Nevertheless, it is not a guarantee that an equally partitioned network will give equal work to all threads. Therefore, future improvements for the simulator should include dynamic partitioning of the network among the threads. This can be implemented by changing the current loop based iteration of network components into a work queue based organization. A work queue can be created for each network partition and each queue entry record the network component to be operated on and the network cycle associated with the operation. After operating on a component, the component is reinserted into the queue and the associated cycle is incremented. Under this organization, threads that have stalled ahead can pull work from other threads' work queues and help to finish the work in other network partitions.

>>>>

\subsection{Static Network Partition}
The current implementation divides the network into equal partitions among the threads and seeks to minimize computation. This organization works well when the traffic pattern is balanced across the partitions. However, hotspot traffic patterns which concentrate the packets into a particular section of the network will cause imbalances in the amount of work handled by each thread. Even with the shared channel semaphore and the removal of the cycle barrier, the lightly loaded threads will  always be stalled waiting for the hotspot threads. Fortunately, common network traffic patterns are balanced and network routing algorithm are usually designed to reduce hotspots and bottlenecks in the network. Nevertheless, it is not a guarantee that an equally partitioned network will give equal work to all threads. Therefore, future improvements for the simulator should include dynamic partitioning of the network among the threads. This can be implemented by changing the current loop based iteration of network components into a work queue based organization. A work queue can be created for each network partition and each queue entry record the network component to be operated on and the network cycle associated with the operation. After operating on a component, the component is reinserted into the queue and the associated cycle is incremented. Under this organization, threads that have stalled ahead can pull work from other threads' work queues and help to finish the work in other network partitions.

### 2009-03-16 doc/paper.tex
I am good with it
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1158 ce7aba0d-9f81-40a6-a495-cecae84b0515

advantage of new multicore processors.  This paper demonstrates how

@@

parameter spaces and consider more exotic configurations.}

@@

increased importance of interconnect network design.  As the number of

@@

of data. The traditional bus architecture is no longer the appropriate medium of communication for large number of cores. Even more recently, single crossbar communication hubs will also not be sufficient in the future since the complexity of a single crossbar grows quadratically with the number of cores. Interconnection network can provide the scalable and efficient communication

@@

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure \ref{lfqueue}.
\begin{figure}[h]
\centering
\includegraphics[width=3in, height=2.25in]{lfqueue.eps}
\caption{This figure demonstrates the layout of the objects associated with a lock-free queue of length 2 (the sentinel value does not count to towards the length of the list).  Note that the pointer to the next element in the list, does not point directly to next list node, but instead to the wrapper AtomicReference class.  Also note that the sentinel is not an actual maintained pointer, but will always point to the same AtomicReference as the Head Node.  It is included in this diagram for clarity.}
\label{lfqueue}
\end{figure}
We initially implemented this AtomicReference object with a PThread
lock.  While this provided a finer granularity of access to shared
objects than previously, we realized that the overhead in the number of
locks was tremendous.  Instead we relied upon an extension to GCC 4.1.2
that provided an atomic \_\_compare\_and\_exchange operation.
This operation was significantly less costly in terms of synchronization
overhead and was also much faster.\\
~\\
After testing our lock-free queues, we replaced all of
the queues in the simulator with our lock-free queue implementation.  While this
proved fairly straight forward, we didn't initially account for the high
overhead that the lock-free queues themselves created.  The lock-free
queues involve the creation of multiple objects and accessing the
lock-free queues involves chasing many pointers through these structures
as well as rearranging many pointers.  In some cases this overhead proved
to be extreme.  In order to remove it, we created a common interface that
was shared between the lock-free queues and the STL queue.  This then
allowed us to create dynamic instances of either a shared queue or an STL
queue depending on whether the channel was shared or not.  This significantly
improved our performance as we will demonstrate in section \ref{results}.

\subsubsection{Guaranteeing Correctness Without a Barrier}
After adding the lock-free queues to our implementation, we realized that
they would provide little benefit if the barrier were not also removed to
allow threads to run ahead.  In order to remove the barrier and still ensure correctness of the simulator, we had to add
an additional piece of data to each shared queue: a semaphore.  This
semaphore would be used in the following way: anytime a write operation
occurred to the queue, an up operation was performed.  Similarly, anytime
a pop operation occurred to the queue, a down operation would have to be
performed first.  We also modified the routers such that for every cycle
of simulation, they would perform at least one up operation on every
channel that they shared with another thread even if they didn't actually
write any flits to that channel.  Note that a down operation is not able
to proceed if the semaphore equals zero, implying that no router can ever
run ahead of any of its input queues.  By adding this feature to our
lock-free queues we were able to guarantee correctness in the simulator
without a barrier operation.  This greatly increased the amount of
concurrency present in the simulation by allowing different threads to be
simulating different cycles simultaneously.

\subsection{Fusing Loops}
The last operation that we performed on our code was to fuse the inner loops in figure~\ref{fig:runtime} together.  The initial simulator implementation was written to make it
easy to understand the code and not for performance.  We moved all these operations into a single
'for' loop in order to help improve some of the temporal locality of the
code.  Not surprisingly this operation allowed the system to take
advantage of the cache hierarchy and provided additional performance gain.

\section{Results \label{results}}
Having completed a parallel implementation of our network simulator, our
next step was to analyze the performance of the simulator to determine how
much speedup it achieved.  We began first by analyzing the performance of
the different synchronization mechanisms to ensure that they held with our
hypotheses about which method permitted the most concurrency.  We then analyzed
our results on a series of architectures to determine the scalability of
the simulator as well as any potential bottlenecks.

@@

\subsection{Locating Scalability Bottlenecks}
Having observed the somewhat poor performance of our application we then
set about attempting to determine the reason for such poor performance.
Our initial hypothesis was that application was memory bound based on the
fact that there is a significant amount of data being processed in our
simulator.  In addition to this, the fact that PThreads is a shared memory
programming model implies that if the working set cannot fit inside the L2
cache, then there will be a significant amount of data being moved both to
and from main memory.\\
~\\
To test our hypothesis that the simulator was memory bound, we ran the
simulator again on the Hotbox cluster, but this time ran each thread on a
different node.  This allowed each thread to have access to the entire
memory bandwidth of its node, and in effect, thereby increase the total
amount of memory bandwidth available to the entire application.  The
results of this experiment can be seen in Figure \ref{smp}.\\
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{smp.eps}
\caption{\label{smp}This figure shows the performance of the simulator on
up to 16 threads on a simulated SMP machine on the Hotbox cluster.  The speedup is relative to
the sequential simulator.}
\end{figure}
From the result of this experiment we were able to see that bandwidth was
not the primary issue.  This is due to the fact that the performance of
the simulator remained poor and was even worse than the CMP version up to
16 threads.  In addition to revealing that the poor performance was not
due to memory bandwidth, it also revealed that communication was playing a
large role in the performance as the increased cost of communicating via
ethernet between nodes caused performance to be even worse than the CMP
configuration.\\
~\\
After realizing that our program was not memory bound, we wanted to see if
the size of the problem had any effect on the speedup.  We tested this by
plotting the speedup of a four threaded run on a given network size over the
sequential time for simulating the same size network on the Hotbox
cluster.  The results can be seen in Figure \ref{probsize}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{probsize.eps}
\caption{\label{probsize}This figure shows the speedup versus network size of a
four threaded run of the parallel simulator on the Hotbox cluster.}
\end{figure}
From this figure we could see that essentially the speedup was independent
of problem size.  This meant that regardless of the size of the network
simulated, the four-threaded version ultimately was constrained by the
same effect as the sequential version.  The fact that the speedup
attained was significantly less than a linear speedup of 4 indicated that
the threads were not compute bound.  Instead we hypothesized that false
sharing was causing the four threads to thrash each others' caches.\\
~\\
To confirm our diagnosis of false sharing, we decided to run the simulator
on the both the Niagara T2 and T2+ machines.  Since these machines were
better designed to deal with long latency loads and stores, our hope was
that they would be able to mitigate the effects of false sharing to some
extent.  The results from this experiment can be seen in figure
\ref{niagara}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{niagara.eps}
\caption{\label{niagara}This figure shows the speedup of the parallel simulator
up to 64 threads on the Niagara T2 and T2+ machines.}
\end{figure}
We noticed that in this experiment, both the T2 and T2+ were able to
achieve some speedup to a certain number of threads, but then performance
began to fall off.  We determined that once the threads' working set sizes
began to interfere with each other in the shared L2 cache is when the
performance begins to degrade and couldn't offer us much insight into  any potential false sharing that was occurring when run on other
hardware platforms.

>>>>

advantage of the new multicore processors.  This paper demonstrates how

@@

parameter spaces and consider a wider range of network configurations.}

@@

increased importance of interconnection network design.  As the number of

@@

of data. The traditional bus architecture is no longer the appropriate medium of communication for large number of cores. Even more recently, single crossbar communication hubs will also not be sufficient in the future since the cost of a single crossbar grows quadratically with the number of cores. Interconnection network can provide the scalable and efficient communication

@@

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure \ref{lfqueue}.\\
~\\
\begin{figure}[h]
\centering
\includegraphics[width=3in, height=2.25in]{lfqueue.eps}
\caption{This figure demonstrates the layout of the objects associated with a lock-free queue of length 2 (the sentinel value does not count to towards the length of the list).  Note that the pointer to the next element in the list, does not point directly to next list node, but instead to the wrapper AtomicReference class.  Also note that the sentinel is not an actual maintained pointer, but will always point to the same AtomicReference as the Head Node.  It is included in this diagram for clarity.}
\label{lfqueue}
\end{figure}

We initially implemented this AtomicReference object with a PThread
lock.  While this provided a finer granularity of access to shared
objects than previously, we realized that the overhead in the number of
locks was tremendous.  Instead we relied upon an extension to GCC 4.1.2
that provided an atomic \_\_compare\_and\_exchange operation.
This operation was significantly less costly in terms of synchronization
overhead and was also much faster.\\
~\\
After testing our lock-free queues, we replaced all of
the queues in the simulator with our lock-free queue implementation.  While this
proved fairly straight forward, we didn't initially account for the high
overhead that the lock-free queues themselves created.  The lock-free
queues involve the creation of multiple objects and accessing the
lock-free queues involves chasing many pointers through these structures
as well as rearranging many pointers.  In some cases this overhead proved
to be extreme.  In order to remove it, we created a common interface that
was shared between the lock-free queues and the STL queue.  This then
allowed us to create dynamic instances of either a shared queue or an STL
queue depending on whether the channel was shared or not.  This significantly
improved our performance as we will demonstrate in section \ref{results}.

\subsubsection{Guaranteeing Correctness Without a Barrier}
After adding the lock-free queues to our implementation, we realized that
they would provide little benefit if the barrier were not also removed to
allow threads to run ahead.  In order to remove the barrier and still ensure correctness of the simulator, we had to add
an additional piece of data to each shared queue: a semaphore.  This
semaphore would be used in the following way: anytime a write operation
occurred to the queue, an up operation was performed.  Similarly, anytime
a pop operation occurred to the queue, a down operation would have to be
performed first. The value of the semaphore is initialized to latency of the channel. We also modified the routers such that for every cycle
of simulation, they would perform at least one up operation on every
channel that they shared with another thread even if they didn't actually
write any flits to that channel.  Note that a down operation is not able
to proceed if the semaphore equals zero, implying that no router can ever
run ahead of any of its input queues.  By adding this feature to our
lock-free queues we were able to guarantee correctness in the simulator
without a barrier operation.  This greatly increased the amount of
concurrency present in the simulation by allowing different threads to be
simulating different cycles simultaneously.

\subsection{Fusing Loops}
The last operation that we performed on our code was to fuse the inner loops in figure~\ref{fig:runtime} together.  The initial simulator implementation was written to make it
easy to understand the code and not for performance.  We moved all these operations into a single loop in order to help improve some of the temporal locality of the
code.  Not surprisingly this operation allowed the system to take
advantage of the cache hierarchy and provided additional performance gain.

\section{Results \label{results}}
Having completed a parallel implementation of our network simulator, we first tested its functionality. We generated a series of latency-throughput plot for a dragonfly network running with different traffic patterns. The resulting plot was nearly identical to those of the sequential simulator. On average there was about less than 0.1\% difference between the two versions. The slight different was caused by the randomness of packet injection. \\
~\\
 Our next step was to analyze the performance of the simulator to determine how
much speedup it achieved.  We began first by analyzing the performance of
the different synchronization mechanisms to ensure that they held with our
hypotheses about which method permitted the most concurrency.  We then analyzed
our results on a series of architectures to determine the scalability of
the simulator as well as any potential bottlenecks.

@@

\subsection{Locating Scalability Bottlenecks}
Having observed the somewhat poor performance of our application we then
set about attempting to determine the reason for such poor performance.
Our initial hypothesis was that application was memory bound based on the
fact that there is a significant amount of data being processed in our
simulator.  In addition to this, the fact that PThreads is a shared memory
programming model implies that if the working set cannot fit inside the L2
cache, then there will be a significant amount of data being moved both to
and from main memory.\\
The first method of our program analysis was by using Intel's Vtune performance analyzer \cite{vtune}. Vtune provides a thread profiler for characterizing multi-threaded programs as well as access to all performance counters in Intel processors. The comparison between the profile of the parallel code and that of the sequential code did not reveal much insights. Majority of the execution time is still within the routing step during each simulation cycle. The lock free queues account for about 10\% of the runtime, proving that synchronization is a performance problem. One performance counter that did stand out is the data bus utilization. The utilization rate for the sequential code is 35\% while the parallel code with 2 threads was at 50\%. After tallying the number of cache lines transferred on the data bus (the BUS\_TRANS\_BURST.ALL\_AGENT counter) and use the formula provided by \cite{account}, we calculated that the parallel program uses approximated 3GB/s memory bandwidth from the front side bus. The maximum possible bandwidth on the system is 6.4GB/s, this was a further indication to us that simulator is perhaps memory bound. 
To test our hypothesis that the simulator was memory bound, we ran the
simulator again on the Hotbox cluster, but this time ran each thread on a
different node.  This allowed each thread to have access to the entire
memory bandwidth of its node, and in effect, thereby increase the total
amount of memory bandwidth available to the entire application.  The
results of this experiment can be seen in Figure \ref{smp}.\\
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{smp.eps}
\caption{\label{smp}This figure shows the performance of the simulator on
up to 16 threads on a simulated SMP machine on the Hotbox cluster.  The speedup is relative to
the sequential simulator.}
\end{figure}
From the result of this experiment we were able to see that bandwidth was
not the primary issue.  This is due to the fact that the performance of
the simulator remained poor and was even worse than the CMP version up to
16 threads.  In addition to revealing that the poor performance was perhaps not
due to memory bandwidth, it also revealed that communication was playing a
large role in the performance as the increased cost of communicating via
ethernet between nodes caused performance to be even worse than the CMP
configuration.\\
After putting the memory bandwidth hypothesis to doubt, we wanted to see if
the size of the problem had any effect on the speedup.  We tested this by
plotting the speedup of a four threaded run on a given network size over the
sequential time for simulating the same size network on the Hotbox
cluster.  The results can be seen in Figure \ref{probsize}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{probsize.eps}
\caption{\label{probsize}This figure shows the speedup versus network size of a
four threaded run of the parallel simulator on the Hotbox cluster.}
\end{figure}
From this figure we could see that essentially the speedup was independent
of problem size.  This meant that regardless of the size of the network
simulated, the four-threaded version ultimately was constrained by the
same effect as the sequential version.  The fact that the speedup
attained was significantly less than a linear speedup of 4 indicated that
the threads were not compute bound.  Instead we hypothesized that false
sharing was causing the four threads to thrash each others' caches.\\
To confirm our diagnosis of false sharing, we decided to run the simulator
on the both the Niagara T2 and T2+ machines.  Since these machines were
better designed to deal with long latency loads and stores, our hope was
that they would be able to mitigate the effects of false sharing to some
extent.  The results from this experiment can be seen in figure
\ref{niagara}.
\begin{figure}[h]
\centering
\includegraphics[height=3in,width=3in]{niagara.eps}
\caption{\label{niagara}This figure shows the speedup of the parallel simulator
up to 64 threads on the Niagara T2 and T2+ machines.}
\end{figure}
We noticed that in this experiment, both the T2 and T2+ were able to
achieve some speedup to a certain number of threads, but then performance
began to fall off.  We determined that once the threads' working set sizes
began to interfere with each other in the shared L2 cache is when the
performance begins to degrade and couldn't offer us much insight into  any potential false sharing that was occurring when run on other
hardware platforms.

### 2009-03-16 doc/sim.tex
I am good with it
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1158 ce7aba0d-9f81-40a6-a495-cecae84b0515

Booksim is written in C++ and dynamically allocates a hierarchy of objects to represent the user specified network. Figure~\ref{fig:simulator} shows the anatomy of the simulator at runtime. At the core of every network simulation is a collection of router and channel objects which forms the framework of a network. Each channel acts as a FIFO with a reader and a writer router object. The router objects contain all the data structure and operations necessary to route packets through the network. Similar to networks in real life, routers are mostly isolated from each other and can only communicate through channels or simple global signals. As a result, the operations of individual routers are almost completely independent of other routers in the network. Surrounding the core network framework is the traffic manager which acts as an interface between the user tests and the underlying network. It handles the injection and ejection of traffic as well as gathering simulations statistics for the user. \\

@@

The simulator was originally designed for studying numerous aspects of small to medium sized interconnection networks (less than 100 network nodes). Nearly every aspect of the network is customizable and has been modeled in detail. Performance has not been a concern. However, the group's recent project involves simulating the dragonfly network \cite{cdragon}, which can easily grow to thousands or tens of thousands of nodes. Due to the organization of the simulator, the simulation runtime grows linearly with the number of network nodes and simulation time. A typical simulation of the dragonfly network involves running a networks with 10K nodes for at least 10K cycles to capture the network's steady state. Under the current simulator, this takes about 25 minutes. To generate an accurate latency-throughput plot of the network requires about 10 simulations. Furthermore, to see the effect of changes to a particular aspect of the network requires generating a latency-throughput plot each time. Such long simulation time increase the turn around time for research and limits the number of experiments that can be run reasonably. Currently, the group utilizes program level parallelism by running multiple simulations simultaneously on the cluster. But they are still limited by the runtime of each simulation.

>>>>

Booksim is written in C++ and dynamically allocates a hierarchy of objects to represent the user specified network. Figure~\ref{fig:simulator} shows the anatomy of the simulator at runtime. At the core of every network simulation is a collection of router and channel objects which forms the framework of a network. Each channel acts as a FIFO with a reader and a writer router object. The router objects contain all the data structure and operations necessary to route packets through the network. Similar to networks in real life, routers are mostly isolated from each other and can only communicate through the channels or simple global signals. As a result, the operations of individual routers are almost completely independent of other routers in the network. Surrounding the core network framework is the traffic manager which acts as an interface between the user tests and the underlying network. It handles the injection and ejection of traffic as well as gathering simulations statistics for the user. \\

@@

The simulator was originally designed for studying numerous aspects of small to medium sized interconnection networks (less than 100 network nodes). Nearly every aspect of the network is customizable and has been modeled in detail. Performance has not been a concern. However, the group's recent project involves simulating the dragonfly network \cite{cdragon}, which can easily grow to thousands or tens of thousands of nodes. Due to the organization of the simulator, the simulation runtime grows linearly with the number of network nodes and the number of simulated cycles. A typical simulation of the dragonfly network involves running a networks with 10K nodes for at least 10K cycles to capture the network's steady state. Under the current simulator, this takes about 25 minutes. To generate an accurate latency-throughput plot of the network requires about 10 simulations. Furthermore, to see the effect of changes to a particular aspect of the network requires generating a latency-throughput plot each time. Such long simulation time increase the turn around time for research and limits the number of experiments that can be run reasonably. Currently, the group utilizes program level parallelism by running multiple simulations simultaneously on a cluster. But they are still limited by the runtime of each simulation.
